---
title: "data_wrangling"
format: html
---

```{r}
library(tidyverse)
library(dplyr)
library(sf)
library(openxlsx)
library(janitor)
library(tmap)
library(terra)
library(exactextractr)
library(arrow)
library(sfarrow)
library(geoarrow)
```

## Load Data

Load **Energy Data** & Filter for TEXAS

```{r}
utility_TX <- read.xlsx(here::here("data", "raw", "eia8602024", "1___Utility_Y2024.xlsx"), startRow = 2) %>% clean_names() %>% filter(state == "TX")

plant_TX <- read.xlsx(here::here("data", "raw", "eia8602024", "2___Plant_Y2024.xlsx"), startRow = 2) %>% clean_names() %>% filter(state == "TX") %>% rename(cnty_name = county)

wind_energy_TX <- read.xlsx(here::here("data", "raw", "eia8602024", "3_2_Wind_Y2024.xlsx"), startRow = 2) %>% clean_names() %>% filter(state == "TX") %>% rename(cnty_name = county)

solar_energy_TX <- read.xlsx(here::here("data", "raw", "eia8602024", "3_3_Solar_Y2024.xlsx"), startRow = 2) %>% clean_names() %>% filter(state == "TX") %>% rename(cnty_name = county)

storge_TX <- read.xlsx(here::here("data", "raw", "eia8602024", "3_4_Energy_Storage_Y2024.xlsx"), startRow = 2) %>% clean_names() %>% filter(state == "TX") %>% rename(cnty_name = county)

generator_TX <- read.xlsx(here::here("data", "raw", "eia8602024", "3_1_Generator_Y2024.xlsx"), startRow = 2) %>% clean_names() %>% filter(state == "TX") %>% rename(cnty_name = county)

multifuel_TX <- read.xlsx(here::here("data", "raw", "eia8602024", "3_5_Multifuel_Y2024.xlsx"), startRow = 2) %>% clean_names() %>% filter(state == "TX") %>% rename(cnty_name = county)

```

Load Texas **Socio-demographic**

```{r}
#st_layers(here::here("data", "raw", "ACS_2019_5YR_TRACT_48_TEXAS.gdb"))

# Geom 
geom_texas <- st_read(here::here("data", "raw", "ACS_2019_5YR_TRACT_48_TEXAS.gdb"), layer = "ACS_2019_5YR_TRACT_48_TEXAS")%>% clean_names()

# Socio-demo info 
income_texas <- st_read(here::here("data", "raw", "ACS_2019_5YR_TRACT_48_TEXAS.gdb"), layer = "X19_INCOME")%>% clean_names()
hispanic_texas <- st_read(here::here("data", "raw", "ACS_2019_5YR_TRACT_48_TEXAS.gdb"), layer = "X03_HISPANIC_OR_LATINO_ORIGIN")%>% clean_names()
#race_texas <- st_read(here::here("data", "raw", "ACS_2019_5YR_TRACT_48_TEXAS.gdb"), layer = "X02_RACE")%>% clean_names()
#poverty_texas <- st_read(here::here("data", "raw", "ACS_2019_5YR_TRACT_48_TEXAS.gdb"), layer = "X17_POVERTY")%>% clean_names()

# Meta data 
metadata_socio_texas <- st_read(here::here("data", "raw", "ACS_2019_5YR_TRACT_48_TEXAS.gdb"), layer = "TRACT_METADATA_2019")
```

Load Texas counties

```{r}
#texas_counties_boundaries <- st_read(here::here("data", "raw", "Texas_County_Boundaries"))%>% clean_names() # Not what I want 

texas_counties <- st_read(here::here("data", "raw", "tl_2024_us_county"))%>% clean_names() %>% 
  filter(statefp == '48') %>% # THIS IS WHAT I WANT - Texas
  rename(cnty_name = name) # Rename county name colunm to `cnty_name`
```

Load Texas **environmental data**

```{r}
# GHI data 
sunshine <- terra::rast(here::here("data", "raw", "nsrdbv3_ghi", "Annual GHI", "nsrdb3_ghi.tif"))%>% clean_names()

# Wind data
# Numbers refer to the height above ground (in meters) the wind speed measures were taken 
wind_envir80 <- terra::rast(here::here("data", "raw", "us-wind-data", "wtk_conus_80m_mean_masked.tif"))%>% clean_names()
wind_envir100 <- terra::rast(here::here("data", "raw", "us-wind-data", "wtk_conus_100m_mean_masked.tif"))%>% clean_names()
wind_envir120 <- terra::rast(here::here("data", "raw", "us-wind-data", "wtk_conus_120m_mean_masked.tif"))%>% clean_names()
wind_envir140 <- terra::rast(here::here("data", "raw", "us-wind-data", "wtk_conus_140m_mean_masked.tif"))%>% clean_names()

```

**The wind data**

This data has wind speed measured at a different height above ground (in meters), ranging from 10 to 200m. Taller = faster winds because surface friction slows wind near the ground.

10m — Surface weather, not useful for energy.

80–100m — Onshore wind turbine hub height, most common for wind energy analysis.

120–140m — Modern large turbines (increasingly common in West Texas).

200m — Offshore or research purposes.

# Clean

## Socio-Economic Data

Making socio-demographic data frames spatial by joining the socio-demographic data frames with `geom_texas`. Both data sets are different layers from the same ACS geodatabase and, thus, correspond to each other cleanly.

```{r}
# Make spatial with county names 
income_texas_sf <- dplyr::left_join(geom_texas, income_texas, by = c("geoid_data"= "geoid")) 
hispanic_texas_sf <- dplyr::left_join(geom_texas, hispanic_texas, by = c("geoid_data"= "geoid"))

# Remove non spatial dfs 
rm(list = c('income_texas', 'race_texas'))

# Remove spatial dfs 
#rm(list = c('income_texas_sf', 'race_texas_sf', 'poverty_texas_sf', 'age_sex_texas_sf'))
```

Creating an Income data frame

```{r}
income_filter <- income_texas_sf %>% 
  # Filter for neccessary columns and rename them
  select(c(countyfp, statefp, tractce, geoid, b19013e1)) %>% 
  rename(avg_income = b19013e1) %>% 
  st_drop_geometry() %>% 
  # Join with texas_counties - which has county names and geoids 
  left_join(texas_counties, by = "countyfp") %>% 
  select(-c(statefp.y, geoid.y, classfp, mtfcc, csafp, cbsafp, metdivfp, funcstat, aland, awater)) %>% 
  st_drop_geometry() %>% 
  rename(geoid = geoid.x)

# income_texas_sf %>% filter(is.na(b19013e1)) # Look at 63 income NA values 

# Create new column of income bins 
income_filter <- income_filter %>% 
  mutate(avg_income_bin = case_when(
    avg_income < 25000 ~ "Under 25K", 
    avg_income < 50000 ~ "25K -50K", 
    avg_income < 100000 ~ "50K - 100K", 
    avg_income < 150000 ~ "100K - 150K", 
    avg_income > 150000 ~ "150K+" 
    ))

```

Creating a Race dataf rame

```{r}
race_filter <- hispanic_texas_sf %>% 
  # Select neccessary columns and rename them 
  select(countyfp, tractce, geoid, b03002e1, b03002e3, b03002e4, b03002e5, b03002e6, b03002e7, b03002e8, b03002e9, b03002e12) %>% 
  rename(
    pop_total = b03002e1,
    white_total = b03002e3,
    black_aa_total = b03002e4,
    ai_an_total= b03002e5, # American Indian and Alaska Native alone
    asian_total = b03002e6, 
    ah_pi_total = b03002e7, # Native Hawaiian and Other Pacific Islander alone
    other_total = b03002e8, # Some other race alone 
    plus2_total = b03002e9,
    hispanic_total = b03002e12) %>% 
  # Join with texas_counties - which has county names and geoids 
  st_drop_geometry() %>% 
  left_join(texas_counties, by = "countyfp") %>% 
  select(-c(geoid.y, classfp, mtfcc, csafp, cbsafp, metdivfp, funcstat, aland, awater)) %>% 
  st_drop_geometry() %>% 
  rename(geoid = geoid.x)


# Chnage race to percentages of total 
count_cols <- c("white_total", "black_aa_total", "ai_an_total", "asian_total", "ah_pi_total", "other_total", "plus2_total", "hispanic_total")
race_filter[, count_cols] <- race_filter[, count_cols] / race_filter$pop_total

```

## Sunshine Data

Crop/Mask to TX

```{r}
# Make CRS match 
texas_counties_proj <- st_transform(texas_counties, crs(sunshine))

# Crop and mask to TEXAS 
sunshine_masked <- crop(sunshine, texas_counties_proj)
sunshine_masked <- terra::mask(sunshine_masked, texas_counties_proj)
```

Turn rasters into county polygons values

```{r}
# Extract rasters into county polygons 
texas_counties$ghi <- exact_extract(sunshine_masked, texas_counties, 'mean')

# Make sunshine dataframe 
sunshine_df <- data.frame(countyfp =texas_counties$countyfp, 
           cnty_name = texas_counties$cnty_name,
           namelsad = texas_counties$namelsad,
           ghi = texas_counties$ghi 
           #geometry = texas_counties$geometry
           )

sunshine_sf <- data.frame(countyfp =texas_counties$countyfp, 
           cnty_name = texas_counties$cnty_name,
           namelsad = texas_counties$namelsad,
           ghi = texas_counties$ghi, 
           geometry = texas_counties$geometry
           )

sunshine_sf <- st_as_sf(sunshine_sf) # needed this for a 

```

## Wind Data

Crop and mask to Texas

```{r}
# Make CRS match 
texas_counties_proj <- st_transform(texas_counties, crs(wind_envir80))

# Crop and mask to TEXAS 
# 80m 
wind_envir80_masked <- crop(wind_envir80, texas_counties_proj)
wind_envir80_masked <- terra::mask(wind_envir80_masked, texas_counties_proj)
# 100m 
wind_envir100_masked <- crop(wind_envir100, texas_counties_proj)
wind_envir100_masked <- terra::mask(wind_envir100_masked, texas_counties_proj)
# 120m 
wind_envir120_masked <- crop(wind_envir120, texas_counties_proj)
wind_envir120_masked <- terra::mask(wind_envir120_masked, texas_counties_proj)
# 140m 
wind_envir140_masked <- crop(wind_envir140, texas_counties_proj)
wind_envir140_masked <- terra::mask(wind_envir140_masked, texas_counties_proj)

```

Turn rasters into county polygons values

```{r}
texas_counties$wind_80m <- exact_extract(wind_envir80_masked, texas_counties, 'mean')
texas_counties$wind_100m <- exact_extract(wind_envir100_masked, texas_counties, 'mean')
texas_counties$wind_120m <- exact_extract(wind_envir120_masked, texas_counties, 'mean')
texas_counties$wind_140m <- exact_extract(wind_envir140_masked, texas_counties, 'mean')

# Make sunshine dataframe 
wind_texas <- data.frame(
  countyfp = texas_counties$countyfp,
  cnty_name = texas_counties$cnty_name,
  namelsad = texas_counties$namelsad,
  wind_80m = texas_counties$wind_80m, 
  wind_100m = texas_counties$wind_100m, 
  wind_120m = texas_counties$wind_120m, 
  wind_140m = texas_counties$wind_140m 
)

# Make sunshine dataframe 
wind_texas_sf <- data.frame(
  countyfp = texas_counties$countyfp,
  cnty_name = texas_counties$cnty_name,
  namelsad = texas_counties$namelsad,
  wind_80m = texas_counties$wind_80m, 
  wind_100m = texas_counties$wind_100m, 
  wind_120m = texas_counties$wind_120m, 
  wind_140m = texas_counties$wind_140m,
  geometry = texas_counties$geometry
)

wind_texas_sf <- st_as_sf(wind_texas_sf)

```

Remove environmental factors from `texas_counties`

```{r}
texas_counties <- texas_counties %>% 
  select(-c(wind_80m, wind_100m, wind_120m, wind_140m))

```

## Energy Data

### Plant

Clean, fix NAs, and make spatial

```{r}
# Change lat and long to numeric 
plant_TX <- plant_TX %>% 
  mutate(
    latitude = as.numeric(latitude), 
    longitude = as.numeric(longitude)
  )

# Are there are any NAs? - yes 
plant_TX %>% 
  filter(is.na(longitude) | is.na(latitude)) 

# Add lat and long to El Paso Electric Co 
plant_TX <- plant_TX %>%
  mutate(
    latitude  = ifelse(is.na(latitude),  31.759269568576,     latitude),
    longitude = ifelse(is.na(longitude), -106.48537808465731, longitude)
  )

# Make spatial 
plant_TX_sf <- st_as_sf(plant_TX, coords = c('longitude', 'latitude'), crs = st_crs(geom_texas))

st_crs(plant_TX_sf) == st_crs(geom_texas) # crs check 

# JOIN - Add geoid
plant_TX_sf <- st_join(plant_TX_sf, geom_texas)
```

### Generator

**Renewables:**

| Energy                  | Abbreviation            |
|-------------------------|-------------------------|
| Solar                   | SUN                     |
| Wind                    | WND                     |
| Hydroelectric           | WAT                     |
| Geothermal              | GEO                     |
| Biomass / Wood / Biogas | WDS, WDL, OBG, OBL, OBS |
| Landfill Gas            | LFG                     |

**Non-renewables:**

| Energy          | Abbreviation       |
|-----------------|--------------------|
| Natural Gas     | NG                 |
| Coal Types      | COL, BIT, SUB, LIG |
| Oil / Petroleum | DFO, RFO, JF, KER  |
| Nuclear         | NUC                |
| Other Gas       | OG                 |

Make a new column about fuel types

```{r}
# Add a column for fuel type 
renewables <- c("SUN","WND","WAT","GEO","WDS","WDL","OBG","OBL","OBS","LFG")

generator_TX <- generator_TX %>% 
  mutate(fuel_category = case_when(
           energy_source_1 %in% c("SUN") ~ "Solar",
           energy_source_1 %in% c("WND") ~ "Wind",
           energy_source_1 %in% c("WAT") ~ "Hydro",
           energy_source_1 %in% c("NG","OG") ~ "Natural gas",
           energy_source_1 %in% c("COL","BIT","SUB","LIG") ~ "Coal",
           energy_source_1 %in% c("NUC") ~ "Nuclear",
           energy_source_1 %in% renewables  ~ "Other renewable",
           TRUE  ~ "Other" 
  ))
  
```

## Save Data Frames

Non-spatial dfs

```{r}
# Socioeconomic 
write_parquet(x = race_filter, sink = "data/processed/race_texas.parquet")

write_parquet(x = income_filter, sink = "data/processed/income_texas.parquet")

# enviornmental
write_parquet(x = sunshine_df, sink = "data/processed/sunshine_texas.parquet")

write_parquet(x = wind_texas, sink = "data/processed/wind_texas.parquet")

# energy 
write_parquet(x = generator_TX, sink = "data/processed/generator_texas.parquet")
```

Spatial dfs

```{r}
# environemental
st_write_parquet(obj = sunshine_sf, dsn = "data/processed/sunshine_texas_sf.parquet", 
                 geometry_encoding = "WKB")

st_write_parquet(obj = wind_texas_sf, dsn = "data/processed/wind_texas_sf.parquet", 
                 geometry_encoding = "WKB")

# energy 
st_write_parquet(obj = plant_TX_sf, dsn = "data/processed/plant_texas_sf.parquet", 
                 geometry_encoding = "WKB")

```

---

Duckdb vs R's spatial stuff is incompatible.

**Problem:** `sfarrow` is writing version `0.1.0` metadata but DuckDB expects `1.0.0`.

Writing geometries as WKT text column in R:

```{r}
sunshine_sf_edit <- sunshine_sf %>% 
  mutate(geom = st_as_text(geometry)) %>% 
  st_drop_geometry()

wind_sf_edit <- wind_texas_sf %>% 
  mutate(geom = st_as_text(geometry)) %>% 
  st_drop_geometry()

plant_sf_edit <- plant_TX_sf %>% 
  mutate(geom = st_as_text(geometry)) %>% 
  st_drop_geometry()
```

Saving the spatial editted dfs

```{r}
write_parquet(x = sunshine_sf_edit, sink = "data/processed/sunshine_texas_sf_v2.parquet")

write_parquet(x = wind_sf_edit, sink = "data/processed/sunshine_wind_sf_v2.parquet")

write_parquet(x = plant_sf_edit, sink = "data/processed/plant_texas_sf_v2.parquet")
```
