---
title: "enviromental_factors_viz"
format: html
---
```{r}
# For database queries 
library(tidyverse)
library(DBI)
library(dbplyr)
library(duckdb)

# For plots 
library(tidyverse)
library(dplyr)
library(sf)
```


First, need to connect to the database. 
```{r}

# Connection
conn <- DBI::dbConnect(duckdb::duckdb(),
                       dbdir = "TX_energy_grid.db")

# Tables 
DBI::dbListTables(conn)

```


## **QUESTION 1:** ENVIR ALIGNMENT

**Do Texas communities leverage local environmental resources in their energy generation mix?**

---

### Q1.A: Is solar energy adoption associated with solar irradiance levels across Texas communities?

Query for solar energy and sunshine patterns
```{r}
# SQL Query 
solar_energy_texas <- DBI::dbGetQuery(
  conn,
  "SELECT * FROM generator_texas 
    JOIN plant_texas USING (plant_code)
    JOIN sunshine_texas USING (countyfp)
    WHERE fuel_category LIKE 'Solar'"
) 

```

Create a line plot for sunshine and solar energy 
```{r}
solar_energy_subset <- solar_energy_texas %>% 
  select(nameplate_capacity_mw, ghi)

# Plot
ggplot(solar_energy_subset) + 
  geom_point(aes(y = nameplate_capacity_mw, 
                 x = ghi), 
             alpha = 0.8, 
             color = "#C5A100") + 
  geom_smooth(aes(y = nameplate_capacity_mw, 
                 x = ghi), 
              color = "firebrick",
              method = lm, se = FALSE) + 
  theme_classic() + 
  labs(x = "Global Horizontal Irradiance (GHI) (W/m²) ", 
       y = "Solar Energy (mw)", 
       title = "Sunshine (GHI) & Solar Energy Generation ")

```


Query for Avg solar energy (mw) per county 
```{r}
# SQL Query 
solar_energy_county <- DBI::dbGetQuery(
  conn,
  "SELECT countyfp, AVG(nameplate_capacity_mw) AS avg_mw FROM generator_texas 
    JOIN plant_texas USING (plant_code)
    JOIN sunshine_texas USING (countyfp)
    WHERE fuel_category LIKE 'Solar'
    GROUP BY countyfp"
)

# Get sunshine dataframe and revive geoms (make spatial again)
sunshine_texas <- tbl(conn, "sunshine_texas") %>% 
  collect() %>% 
  mutate(geom = st_as_sfc(geom)) %>% 
  st_as_sf()

```

Create map of county sunshine and solar energy mw 
```{r}
# Merge 
solar_sf <- sunshine_texas %>%  # suppose to have geometry 
  left_join(solar_energy_county, by = "countyfp")

# Make centeriod 
solar_sf <- st_as_sf(solar_sf)
solar_centroid <- st_centroid(solar_sf)

# Plot 
ggplot() + 
  geom_sf(data = solar_sf,
          mapping = aes(fill = ghi)) + 
  geom_sf(data = filter(solar_centroid, !is.na(avg_mw)), 
          mapping = aes(size = avg_mw), 
          shape = 21, 
          color = "white",
          fill = "black",
          alpha = 0.6) + 
  scale_fill_distiller(palette = "YlOrRd", direction = 1, name = "Solar Irradiance (GHI)") + 
  theme_void()
```


### Q1.B: Is wind energy capacity associated with wind resource availability across Texas communities?

Query for wind energy and wind speed patterns
```{r}
# SQL Query 
wind_energy_texas <- DBI::dbGetQuery(
  conn,
  "SELECT * FROM generator_texas 
    JOIN plant_texas USING (plant_code)
    JOIN wind_texas USING (countyfp)
    WHERE fuel_category LIKE 'Wind'"
)

```

Create a line plot for sunshine and solar energy 
```{r}
wind_energy_subset <- wind_energy_texas %>% 
  select(c(wind_100m, wind_120m, nameplate_capacity_mw))

ggplot(wind_energy_subset) + 
  geom_point(aes(y = nameplate_capacity_mw, 
                 x = wind_100m), 
             alpha = 0.5, 
             color = "#00A8EA") + 
  geom_smooth(aes(y = nameplate_capacity_mw, 
                 x = wind_100m), 
              color = "firebrick",
              method = lm, se = FALSE) + 
  theme_classic() + 
  labs(x = "Wind Speed (m/sec) at 100m", 
       y = "Wind Energy (mw)", 
       title = "Wind Speeds & Wind Energy Generation ")


ggplot(wind_energy_subset) + 
  geom_point(aes(y = nameplate_capacity_mw, 
                 x = wind_120m), 
             alpha = 0.5, 
             color = "#00A8EA") + 
  geom_smooth(aes(y = nameplate_capacity_mw, 
                 x = wind_120m), 
              color = "firebrick",
              method = lm, se = FALSE) + 
  theme_classic() + 
  labs(x = "Wind Speed (m/sec) at 120m", 
       y = "Wind Energy (mw)", 
       title = "Wind Speeds & Wind Energy Generation ")

```


Query for Avg solar energy (mw) per county 
```{r}
# SQL Query 
wind_energy_county <- DBI::dbGetQuery(
  conn,
  "SELECT countyfp, AVG(nameplate_capacity_mw) AS avg_mw FROM generator_texas 
    JOIN plant_texas USING (plant_code)
    JOIN wind_texas USING (countyfp)
    WHERE fuel_category LIKE 'Wind'
    GROUP BY countyfp;"
)

wind_texas <- tbl(conn, "wind_texas") %>% 
  collect() %>% 
  mutate(geom = st_as_sfc(geom)) %>% 
  st_as_sf()

```


Create map of county wind speeds and wind energy mw 
```{r}
# Merge 
wind_sf <- wind_texas %>%  # suppose to have geometry 
  left_join(wind_energy_county, by = "countyfp")

# Make centeriod 
wind_sf <- st_as_sf(wind_sf)
wind_centroid <- st_centroid(wind_sf)

# Plot 
ggplot() + 
  geom_sf(data = wind_sf,
          mapping = aes(fill = wind_120m)) + 
  geom_sf(data = filter(wind_centroid, !is.na(avg_mw)), 
          mapping = aes(size = avg_mw), 
          shape = 21, 
          color = "white",
          fill = "black",
          alpha = 0.6) + 
  scale_fill_distiller(palette = "Blues", direction = -1, name = "Wind at 120m") + 
  theme_void()

```



# Close databse 
```{r}
# CLOSE DATABASE CONNECTION  - MUST DO 
DBI::dbDisconnect(conn, shutdown = TRUE)

```