---
title: "socioEconomic_viz"
format: html
---

```{r}
# For database queries 
library(tidyverse)
library(DBI)
library(dbplyr)
library(duckdb)

# For plots 
library(tidyverse)
library(dplyr)
library(sf)
```

First, need to connect to the database.

```{r}

# Connection
conn <- DBI::dbConnect(duckdb::duckdb(),
                       dbdir = "TX_energy_grid.db")

# Tables 
DBI::dbListTables(conn)

```

## **QUESTION 2:** ENVIR ALIGNMENT

**How is energy generation distributed across socioeconomically disadvantaged communities in Texas?**

------------------------------------------------------------------------

### Q2.A: How does residential energy source distribution vary across income brackets? 

Query for (1) median income and total energy per geoid and (2) Find median income per geiod 
```{r}
# SQL Query 
# (1) median income and total energy per fuel type by geoid 
income_energy_fuelType <- DBI::dbGetQuery(
  conn,
  "SELECT 
    g.geoid,g.fuel_category, g.total_mw, i.median_avg_income
FROM (
    SELECT geoid, fuel_category, SUM(nameplate_capacity_mw) AS total_mw 
    FROM generator_texas 
    JOIN plant_texas USING (plant_code)
    GROUP BY geoid, fuel_category
) g
JOIN (
    SELECT geoid, MEDIAN(avg_income) AS median_avg_income 
    FROM income_texas
    GROUP BY geoid
) i USING (geoid);"
)


income_energy_fuelType
```

Create and plot Income classes Energ composition  
```{r}
# Add income bin 
income_energy_fuelType <- income_energy_fuelType %>% 
  mutate(income_class = case_when(
    median_avg_income < 25000 ~ "Under 25K", 
    median_avg_income < 50000 ~ "25K -50K", 
    median_avg_income < 100000 ~ "50K - 100K", 
    median_avg_income < 150000 ~ "100K - 150K", 
    median_avg_income > 150000 ~ "150K+" 
  ))

# color palette 
fuel_cat_pal <- c(
  'Coal' = '#BB8900',
  'Hydro' = '#2C549C',
 'Natural gas' = '#B0B0B0', 
  'Nuclear' = '#D04A39',
  'Other' = '#484848',
  'Other renewable' = '#02BA00',
  'Solar' = '#FCDB48',
  'Wind' = '#00A8EA'
)

# Plot 
income_energy_plot <- income_energy_fuelType %>%
  filter(!is.na(income_class)) %>%
  group_by(income_class, fuel_category) %>%
  summarise(TOTAL_mw = sum(total_mw, na.rm = TRUE), 
            .groups = "drop") %>%
  ggplot() +
  geom_col(aes(x = income_class, y = TOTAL_mw, fill = fuel_category), position = "fill") +
  scale_y_continuous(labels = scales::percent) +
  scale_fill_manual(values = fuel_cat_pal) +
  coord_flip() +
  labs(y = "Share of capacity", title = "Census Tracks Income and Energy Composition") +
  theme_light()

income_energy_plot
# Save plot 
ggsave(filename = here::here('figs', 'income_energy_plot.pdf'), plot = income_energy_plot)

```


### Q2.B: How does energy generation composition vary across communities of different racial and ethnic demographics?

Query for the dataframe about texas's racial composition 
```{r}
race_texas <- tbl(conn, "race_texas") %>% 
  collect()
```

Create a new column of the dominate race in each census track and merge with energy data 
```{r}
# New column of the dominate race in each census track  
race_texas <- race_texas %>%
  mutate(dominant_race = case_when(
    white_total    == pmax(white_total, black_aa_total, hispanic_total, asian_total, other_total) ~ "White",
    black_aa_total == pmax(white_total, black_aa_total, hispanic_total, asian_total, other_total) ~ "Black",
    hispanic_total    == pmax(white_total, black_aa_total, hispanic_total, asian_total, other_total) ~ "Hispanic",
    asian_total    == pmax(white_total, black_aa_total, hispanic_total, asian_total, other_total) ~ "Asian",
    TRUE ~ "Other"
  ))

# Select just the energy data from income dataframe above 
fuel_catagories_mw <- income_energy_fuelType %>% 
  select(geoid, fuel_category, total_mw)

# Join fuels and race data 
race_energy <- race_texas %>% 
  left_join(fuel_catagories_mw, by = "geoid") 

```

```{r}
race_energy_plot <- race_energy %>% 
  #filter(!dominant_race %in% c("Other", "Asian")) %>%
  group_by(dominant_race, fuel_category) %>% 
  summarise(TOTAL_mw = sum(total_mw, na.rm = TRUE), 
            .groups = "drop") %>% 
  ggplot() + 
  geom_col(aes(y = dominant_race, x = TOTAL_mw, fill = fuel_category), position = "fill") +
  scale_x_continuous(labels = scales::percent) +
  scale_fill_manual(values = fuel_cat_pal) +
  labs(y = "Share of capacity", 
       title = "Census Tracks Income and Energy Composition") + 
  theme_light()

# save plot
ggsave(filename = here::here('figs', 'race_energy_plot.pdf'), plot = race_energy_plot)

```

# Close Database

```{r}
# CLOSE DATABASE CONNECTION  - MUST DO 
DBI::dbDisconnect(conn, shutdown = TRUE)

```

